On Optimality of Finite Element Integration

### 2015-10-17 TOMS_Paper/assembly.tex
Started incorporating David's suggestions

In this section, we characterize optimality for finite element integration and the loop transformation space that we need to explore to achieve it.

@@

\begin{Def}[Linear loop]
A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is \textit{linear} if 
\begin{enumerate}
\item $i$ appears in the body of $L$ only as an array index, and
\item whenever an array $a$ is indexed by $i$ ($a[i]$), all expressions in which this appears are affine in $a$.
\end{enumerate}
\end{Def}

@@

As shown later, outer product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code. They arise naturally when translating bilinear or linear forms into code. Consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises the element loop $L_e$, the integration loop $L_i$ (a reduction loop), and the outer product loop nest $[L_j, L_k]$ over test and trial functions. The expression \texttt{F} implements the operator.

% (the Laplacian operator in the case of Equation~\ref{eq:quadrature}, but our discussion applies in general).
\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{The typical loop nest implementing a bilinear form.}\label{code:loopnest}\end{figure}

Our aim is a strategy to synthesize optimal loop nests of this form. In particular, we characterize optimality as follows.

>>>>

In this section, we characterize optimality for finite element integration and the transformation space that we need to explore to achieve it.

@@

\begin{Def}[Linear loop]
A loop $L$ defining the iteration space $I$ through the iteration variable $i$, or simply $L_i$, is linear if in its body
\begin{enumerate}
\item $i$ appears only as an array index, and
\item whenever an array $a$ is indexed by $i$ ($a[i]$), all expressions in which this appears are affine in $a$.
\end{enumerate}
\end{Def}

@@

As shown later, outer product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code. We will see that they arise naturally when translating bilinear or linear forms into code.

Two other classes of loops need be characterized.

\begin{Def}[Reduction loop]
A loop $L_i$ is said to be a reduction loop if in its body
\begin{enumerate}
\item $i$ appears only as an array index, and
\item for each augmented assignment statement $S$ (e.g., increments), arrays indexed by $i$ appear only on the right hand side of $S$.
\end{enumerate}
\end{Def}

\begin{Def}[Free space loop]
A loop $L_i$ is said to be a free space loop if its iterations can be executed in any arbitrary order; that is, no dependencies exist across different iterations. 
\end{Def}

Altogether, these definitions lead us to formalize the class of loop nests for which we seek optimality.

\begin{Def}[Finite element integration loop nest]
A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free space loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
\end{Def}

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free space loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions). The expression \texttt{F} implements the operator.

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{The typical loop nest implementing a bilinear form.}\label{code:loopnest}\end{figure}

We emphasize that our aim is a strategy to synthesize optimal loop nests of this specific form. In particular, we characterize optimality as follows.
